Type Evolution in Programming Systems

### 2024-05-29 paper.tex
cleanup

\DeclareRobustCommand{\frameworkbox}[2][gray!15]{
\begin{tcolorbox}[breakable,left=3pt,right=3pt,top=3pt,bottom=3pt,colback=#1,colframe=#1,parbox=false,
  width=\dimexpr\textwidth\relax,enlarge left by=0mm,boxsep=5pt,arc=0pt,enlarge top by=0.5em,%enlarge bottom by=0.0em,
  outer arc=0pt]\setlength{\parskip}{0.5em}#2
\end{tcolorbox}}
\newcommand{\frameworkboxtitle}[1]{\setlength{\parindent}{0em}{\firamedium #1.}\quad}
\newcommand{\openframeworkbox}{
\begin{tcolorbox}[breakable,left=3pt,right=3pt,top=3pt,bottom=3pt,colback=gray!15,colframe=gray!15,parbox=false,
  width=\dimexpr\textwidth\relax,enlarge left by=0mm,boxsep=5pt,arc=0pt,enlarge top by=0.5em,%enlarge bottom by=0.0em,
  outer arc=0pt]\setlength{\parskip}{0.5em}
}
\newcommand{\closeframeworkbox}{
\end{tcolorbox}
}

@@

\textcolor{red}{[JE. Found this snippet in the related work]
Acme Corp has a problem, for spreadsheets don't naturally support relationships between entities.
End-user databases like Airtable~\cite{airtable} and Notion~\cite{notion} support \textit{links}
between tables but do not provide operations that can do an \textit{Extract Entity}. Neither do
the established database products (see Related Work). Acme will need to write code or do a big manual conversion.
}

@@

\openframeworkbox
  \begin{tikzcd}[sep = 6em]
    \textsf{type}_1
    \arrow[r, "\textsf{constrains}"{above}]
    \arrow[d, red, "\textsf{change}_1"{name=C, description}]
    &
    \textsf{data}_1
    \arrow[d, "\textsf{migrate}_1"{name=R, description}]
    \\
    \textsf{type}_2
    \arrow[r, "\textsf{constrains}"{above}]
    \arrow[d, red, "\textsf{change}_2"{name=C2, description}]
    &
    \textsf{data}_2
    \arrow[Rightarrow, from=C, to=R, "\textsf{propagate}"{above}]
    \arrow[d, "\textsf{migrate}_2"{name=R2, description}]
    \\
    \textsf{type}_3
    \arrow[r, "\textsf{constrains}"{above}]
    &
    \textsf{data}_3
    \arrow[Rightarrow, from=C2, to=R2, "\textsf{propagate}"{above}]
  \end{tikzcd}

  \frameworkboxtitle{Primitive Type Transformations}
In this example, \textsf{type} refers to data tables, but the same scenario applies to
objects or JSON values in other programming systems. Our example focuses on permanent \emph{data}
that need to evolve alongside with the type and keep synchronized even in the non-local scenario.
The evolution can be done through a small set of primitive type transformations (illustrated on the
right) with corresponding data transformation for each. The system may either expose the
primitive transformations through a user interface, or infer them from textual edits made by the
programmer, e.g. to a database table schema definition.
\closeframeworkbox

@@

\textcolor{red}{Maybe insert something about methodology in general?
(ref the table challenges paper? although this paper is no longer so much about challenges)}

There has been a lot of research on evolving interdependent layers of software, much of it guided by schema and types. But the vast majority of these techniques involve writing custom code in some specialized language, rather than through immediate action in an interactive programming system. Nevertheless it is important to understand these batch techniques to envision more interactive ones.

>>>>

\DeclareRobustCommand{\frameworkbox}[2][gray!15]{
\begin{tcolorbox}[breakable,left=3pt,right=3pt,top=3pt,bottom=3pt,colback=#1,colframe=#1,parbox=false,
  width=\dimexpr\textwidth\relax,enlarge left by=0mm,boxsep=5pt,arc=0pt,enlarge top by=0.5em,%enlarge bottom by=0.0em,
  outer arc=0pt]\setlength{\parskip}{0.5em}#2
\end{tcolorbox}}
\newcommand{\frameworkboxtitle}[1]{\setlength{\parindent}{0em}{\firamedium #1.}\quad}

@@

@@

\frameworkbox{
\begin{wrapfigure}{r}{9em}
\includegraphics[scale=0.2]{figures/diag-variants.png}
\end{wrapfigure}

\frameworkboxtitle{Primitive Type Transformations}
In this example, \textsf{type} refers to data tables, but the same scenario applies to
objects or JSON values in other programming systems. Our example focuses on permanent \emph{data}
that need to evolve alongside with the type and keep synchronized even in the non-local scenario.
The evolution can be done through a small set of primitive type transformations (illustrated on the
right) with corresponding data transformation for each. The system may either expose the
primitive transformations through a user interface, or infer them from textual edits made by the
programmer, e.g. to a database table schema definition.
}

@@

There has been a lot of research on coordinating the evolution of interdependent layers of software, much of it guided by schema and types. But the vast majority of these techniques involve writing custom code in some specialized language, rather than through immediate action in an interactive programming system. Nevertheless it is important to understand these batch techniques to envision more interactive ones.
